A Transactional Memory System for Parallel Python

### 2014-07-31 talk/dls2014/paper/paper.tex
update benchmark numbers in paper

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language's implementations is the use of
a single global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, applications can not make use of the increasing parallelism in
current hardware.

@@

Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot
of attention but only provide limited forms of parallelism. A parallel
programming model was not part of the initial design of those languages. Thus,
the reference implementations of, e.g., Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads. The use of a single global lock causes several disadvantages,
and as multi-core processors become the de facto standard in platforms
ranging from smart phones to data centres, the issue of parallel
execution must be addressed by these dynamic languages.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on multiple
threads. One approach to this is fine-grained locking instead of a
single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We on the other hand follow the direct approach of using TM
instead of the GIL. There, GIL acquisition and release operations
conceptually directly map to starting and committing transactions.

@@

We performed all benchmarks on a machine with an Intel Core i7-4770
CPU~@3.40GHz (4 cores, 8 threads) with disabled Hyper-Threading and
Turbo Boost for less variation in the results.  There are 16~GiB of memory
available and we ran them under Ubuntu 14.04 with a Linux 3.13.0
kernel. The STM system was compiled with a number of segments $N=4$
and a maximum amount of memory of 1.5~GiB (both are configurable at
compile time).

@@

The STM system detected no conflicts when running this code on 4
threads. For the results in Figure~\ref{fig:scaling}, we
normalised the average runtimes to the time it took on a single
thread. From this we see that there is some additional overhead introduced
by each thread ($12.3\%$ for all 4 threads together). Every thread
adds some overhead because during a commit, there is one more thread
which has to reach a safe point. Additionally, conflict detection
needs to check for conflicts in all concurrently running transactions.

While not ideal, we think that $12.3\%$ is acceptable on four
threads. In terms of throughput, 4 threads have $3.54\times$
more iterations per second than a single thread.

@@

For the following sections we use a set of six small benchmarks
available at~\cite{pypybenchs}. There are, unsurprisingly, not
many threaded applications written in Python that can be used
as a benchmark. Since until now, threading was rarely used
for performance reasons because of the GIL, we mostly collected
small demos and wrote our own benchmarks to evaluate our system:

\begin{itemize}
\item \emph{btree} and \emph{skiplist}, which are both inserting,
  removing, and finding elements in a data structure from multiple
  threads
\item \emph{threadworms}, which simulates worms walking on a grid in
  parallel and checking for collisions with each other
\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
  all perform some independent computations in parallel (embarrassingly
  parallel)
\end{itemize}

We use atomic blocks to easily synchronise access to shared
data structures in the first three benchmarks. For the embarrassingly
parallel benchmarks, we do not need any explicit synchronisation at
all. These atomic blocks are simulated with a single global lock
when running on top of GIL-supported interpreters. With our STM
system, they map to transactions that will execute optimistically
in parallel.

@@

On 4 cores, PyPy using our STM system (\emph{pypy-stm-nojit}) scales
in all benchmarks to a certain degree. It scales best for the
embarrassingly parallel ones ($avg=2.6\times$ speedup over 1 thread)
and a little less for the others ($avg=2.0\times$ speedup over 1
thread). The reason for this difference is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

Looking at the average overhead on a single thread that is induced by
switching from GIL to STM, we see that it is $\approx 43.3\%$. The
maximum in richards is $71\%$. In all benchmarks \emph{pypy-stm-nojit}
beats \emph{pypy-nojit} already on two threads, despite of this
overhead.  The achieved speedup comparing the lowest runtimes of STM
to the single-threaded GIL execution is between $1.14\times$ and
$1.94\times$.

Still, STM rarely beats CPython's \emph{single-thread} performance. However, for
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

We would like to regard enabling the JIT as a simple performance
enhancement, but that is not what happens in reality. First, since the
JIT~\cite{cfbolz09} bases its decision of what to compile on runtime
profiling, running in multiple threads, it may compile different
things in each run because of the non-deterministic thread-scheduling
of the operating system (OS). Second, it is able to perform some
advanced optimisations. Because compilation is already non-deterministic,
so are these optimisations. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.
For these reasons, the following results have to be taken with a grain
of salt.

@@

The results are presented in Figure~\ref{fig:performance-jit}. We see
that the performance gains are much less reliable.  The slowdown
factor for switching from GIL to STM ranges around $1-2.4\times$, and
we beat the GIL's single-thread performance in half of the benchmarks.

We see that generally, the group of embarrassingly parallel benchmarks
scales best. (There is a notable performance stability problem in the
\emph{richards} benchmark. This is a bug we will fix in the future.)
The other three benchmarks scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Right now, because the code runs much more quickly with
the JIT than without, it has the effect of making the transactions
logically longer.  This increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

Overall, PyPy without STM is around $2\times$ slower than CPython.
Enabling its JIT allows it to outperform CPython by a huge margin.
We see the same kind of speedup on PyPy with STM when enabling the
JIT. This means that STM does not generally inhibit the JIT from
optimising the programs execution, which is already a very important
result on its own. We also see that for some
benchmarks, STM is already able to give additional speedups
compared to just the JIT-induced acceleration. This looks very
promising and investigating when this is not the case is the next
logical step.

@@

% Similar STMs:
% \begin{itemize}
% \item FastLane: \cite{warmhoff13}
% \item TML: \cite{spear09}
% \item Virtualizing HTM: \cite{rajwar05}
% \item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
%   detection, otherwise hardware extensions required; assumes most
%   transactions fit HTM capacities (not so true here); COW using page-faults;
%   they assume OS-level access to page-tables (maybe not inherent to their
%   design); eval on simulator; value-based confl detection;

%  (XTM can be
%   implemented either in the OS as part of the virtual memory manager or
%   between underlying TM systems and the OS, like virtual machines;
%   Conflicts for overflowed transactions are tracked at page granularity;
%   XTM-e allows conflict detection at cache line granu-
%   larity, even for overflowed data in virtual memory)
% \item using mmap(): Memory-Mapped Transactions
% \item mem-protected conflict detection: \cite{martin09}
% \end{itemize}

@@

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement scales well and yields an average speedup of $2.6\times$ for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches $2.0\times$.

>>>>

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language's implementations is the use of
a single, global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, applications can not make use of the increasing parallelism in
current hardware.

@@

Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot
of attention but only provide limited forms of parallelism. A parallel
programming model was not part of the initial design of those languages. Thus,
the reference implementations of, e.g., Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads. The use of a single, global lock causes several disadvantages,
and as multi-core processors become the de facto standard in platforms
ranging from smart phones to data centres, the issue of parallel
execution must be addressed by these dynamic languages.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on multiple
threads. One approach to this is fine-grained locking instead of a
single, global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We on the other hand follow the direct approach of using TM
instead of the GIL. There, GIL acquisition and release operations
conceptually directly map to starting and committing transactions.

@@

We performed all benchmarks on a machine with an Intel Core i7-4770
CPU~@3.40GHz (4 cores, 8 threads) with disabled Hyper-Threading and
Turbo Boost for less variation in the results.  There are 16~GiB of memory
available and we ran them under Ubuntu 14.04 with a Linux 3.13.0
kernel. The STM system was compiled with a number of segments $N=4$
and a maximum amount of memory of 1.5~GiB (both configured at
compile time).

@@

The STM system detected no conflicts when running this code on 4
threads. For the results in Figure~\ref{fig:scaling}, we
normalised the average runtimes to the time it took on a single
thread. From this we see that there is some additional overhead introduced
by each thread ($9.1\%$ for all 4 threads together). Every thread
adds some overhead because during a commit, there is one more thread
which has to reach a safe point. Additionally, conflict detection
needs to check for conflicts in all concurrently running transactions.

While not ideal, we think that $9.1\%$ is acceptable on four
threads. In terms of throughput, 4 threads reach $3.67\times$
the iterations per second of a single thread.

@@

For the following sections we use a set of 8 small benchmarks
available at~\cite{pypybenchs}. There are, unsurprisingly, not
many threaded applications written in Python that can be used
as a benchmark. Since until now, threading was rarely used
for performance reasons because of the GIL, we mostly collected
small demos and wrote our own benchmarks to evaluate our system:

\begin{itemize}
\item \emph{btree} and \emph{skiplist}, which are both inserting,
  removing, and finding elements in a data structure from multiple
  threads
\item \emph{threadworms}, which simulates worms walking on a grid in
  parallel and checking for collisions with each other
\item \emph{miller-rabin}, \emph{mandelbrot}, \emph{raytrace},
  \emph{richards}, and \emph{mersenne}, which all perform some mostly
  independent computations in parallel (embarrassingly parallel).
\end{itemize}

We use atomic blocks as the primary mechanism to synchronise access to
shared data structures in the first three benchmarks. For the
embarrassingly parallel benchmarks, we only need explicit
synchronisation in some small isolated part. These atomic blocks are
simulated with a single, global lock when running on top of
GIL-supported interpreters. With our STM system, they map to
transactions that will execute optimistically in parallel.

@@

On 4 cores, PyPy using our STM system (\emph{pypy-stm-nojit}) scales
in all benchmarks to a certain degree. \emph{mersenne} is a bit of a
special case, since it is easily parallelisable but a few of the
parallel computations dominate the runtime and cannot be made faster
easily by parallelisation. This is why the benefit of adding more
threads than 2 is diminishing.

\emph{pypy-stm-nojit} scales best for the embarrassingly parallel
benchmarks ($avg=2.7\times$ speedup over 1 thread) and a little less
for the others ($avg=2.5\times$ speedup over 1 thread). The reason for
this difference is that in the former group there are no real, logical
conflicts -- all threads do independent calculations. STM simply
replaces the GIL in those programs. In the latter group, the threads
work on a common data structure and therefore create much more
conflicts, which limits the scalability. Here we make active use of
the STM-supported atomic blocks to synchronise access to the data
structure. The hope is that the STM system is still able to
parallelise, even if we use the atomic blocks in a coarse-grained
way. While less than the other benchmarks, we indeed see some speedup
going from 1 to 4 threads.

Looking at the average overhead on a single thread that is induced by
switching from GIL to STM, we see that it is $\approx 40.1\%$. The
maximum in richards is $65.5\%$. In all benchmarks \emph{pypy-stm-nojit}
beats \emph{pypy-nojit} already on two threads, despite of this
overhead.  The achieved speedup comparing the fastest runtimes of STM
to the single-threaded GIL execution of \emph{pypy-nojit} is between
$1.11\times$ and $2.55\times$.

Still, \emph{pypy-stm-nojit} barely beats CPython's \emph{single-thread}
performance. However, for
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all benchmarks.

@@

We would like to regard enabling the JIT as a simple performance
enhancement, but that is not what happens in reality. First, since the
JIT~\cite{cfbolz09} bases its decision of what to compile on runtime
profiling, running in multiple threads, it may compile different
things in each run because of the non-deterministic thread-scheduling
of the operating system (OS). Second, it is able to perform some
advanced optimisations. Because compilation is already non-deterministic,
so are these optimisations. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.
For these reasons, the following results should be regarded as a
preliminary outlook.

@@

The results are presented in Figure~\ref{fig:performance-jit}. We see
that the performance gains of STM are a bit less reliable with the JIT.
The slowdown factor for switching from GIL to STM ranges around
$1-2.5\times$, and we beat the GIL's single-thread performance in
5 out of 8 benchmarks.

We see that generally, the group of embarrassingly parallel benchmarks
scales best.  The other three benchmarks scale barely or not at all
with the number of threads. The reason for this is likely again the
conflicts in the latter group. Right now, because the code runs much
more quickly with the JIT than without, it has the effect of making
the transactions logically longer.  This increases the likelihood of
conflicts between them and therefore limits scalability even more than
in the no-JIT benchmarks.

Overall, PyPy without STM is around $2\times$ slower than CPython
(\emph{cpython} vs.\ \emph{pypy-nojit}).  Enabling its JIT allows it
to outperform CPython by a huge margin.  We see the same kind of
speedup on PyPy with STM when enabling the JIT. This means that STM
does not generally inhibit the JIT from optimising the programs
execution, which is already a very important result on its own. We
also see that for some benchmarks, STM is already able to give
additional speedups compared to just the JIT-induced
acceleration. This looks very promising and improving this situation
even more is the next step for reaching the goal of parallelising
Python.

@@

@@

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement scales well and yields an average speedup of $2.7\times$ for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches $2.5\times$.
